Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-08 paper/intro.tex

For a \emph{constant} matrix $\mtx{B}$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}. Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification to these extensions. In passing, we note that dynamic trace estimation~\cite{dharangutte-2021-dynamic-trace,woodruff-2022-optimal-query} is an efficient technique for \textcolor{red}{subsequently ??? true? ???} estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm. The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with 
$g_\sigma$ approximating a Dirac delta function.

@@

\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at ea<ch value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices. We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation in \refsec{sec:results}.

\paragraph{Reproducibility} \textcolor{red}{This paragraph looks odd. Is this the wording required by SISC? If not, focus on the reproducibility of the numerical experiments and figures/tables in this statement.}  \input{re-pro-badge.tex}

>>>>

For a \emph{constant} matrix $\mtx{B}$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}. Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification of these extensions. In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for \textcolor{red}{subsequently ??? true? ???} estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm. The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with $g_{\sigma}$ approximating a Dirac delta function.

@@

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at ea<ch value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

\paragraph{Structure.} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices. We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation in \refsec{sec:results}.

\paragraph{Reproducibility.} \textcolor{red}{This paragraph looks odd. Is this the wording required by SISC? If not, focus on the reproducibility of the numerical experiments and figures/tables in this statement.}  \input{re-pro-badge.tex}

### 2024-07-08 paper/re-pro-badge.tex

This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/ce5dd53}{ce5dd53} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-07 at 11:46:28 UTC.

>>>>

This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-TRACE/tree/04376a9}{04376a9} at \url{https://github.com/FMatti/Rand-TRACE} on 2024-07-07 at 15:28:11 UTC.
